---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Data are first loaded into the workspace using load csv command and additional 
processing to change date columns from char into date. The loaded dataset
consists of 17568 rows & 3 columns (interval, date and steps)
```{r}
dataset <- read.csv("activity.csv",header=T, sep=',', stringsAsFactors=F, 
                    comment.char="",na.strings = "NA")
dataset$date <- as.Date(dataset$date)
```

## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day

Total number of steps taken are first calculated for each date - ignoring any 
missing values. Using base plot, a histogram is made from aggregated results.

```{r}
dailysteps <- aggregate(steps ~ date, data=dataset, sum)
hist(dailysteps$steps, 
     main="Histogram of the total number of steps taken each day"
     , xlab = "Number of steps daily", ylab = "Number of Days (Frequency)"
     , col="#0000FF88")
```


2. Calculate and report the mean and median total number of steps taken per day

The daily mean are taken across all all dates and yield the result of 10766.19
steps per day.
```{r}
mean(dailysteps$steps)
```

The daily median are taken across all all dates and yield the result of 10765
steps per day.
```{r}
median(dailysteps$steps)
```

## What is the average daily activity pattern?
1.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis)

The average steps are calculated for each interval across the date. Again the
missing values in the dataset are removed from the calculation. A trendline
for average number of steps taken vs. interval measured are plotted.

```{r}
avgsteps <- aggregate(steps ~ interval, data=dataset, mean, na.rm=T)
plot(x=avgsteps$interval, y=avgsteps$steps, type="l", 
     main="Time Series Plot of the 5-minute Interval", 
     xlab="5-minute time interval",ylab="average number of steps taken")
```


2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?

The interval 835 contains the maximum number of steps when averaged across all 
the days in the dataset
```{r}
avgsteps[avgsteps$steps == max(avgsteps$steps), ]
```

## Inputing missing values
1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)

For a quick check of missing data, the complete.case() from stats was used to 
detect any missing values in the dataset.
```{r}
sum(!complete.cases(dataset))
```

2. Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use the 
mean/median for that day, or the mean for that 5-minute interval, etc.

Missing data in the original dataset would be replaced with the mean for that
5-minute interval. Due to the variation of activities/number of steps taken
throughout the day, this strategy would be able to represent daily trend better
than single values of mean or median.

3. Create a new dataset that is equal to the original dataset but with the 
missing data filled in.

In order to fill in the missing data, first a new dataset "new_ds" were created.
The columns steps would be updated via a conditional statement. When steps 
information are missing, the calculated from average steps for the particulaer
interval would be substituted into the dataset.
```{r}
new_ds <- dataset
new_ds$steps <- ifelse(is.na(new_ds$steps), 
                        avgsteps$steps[match(new_ds$interval, avgsteps$interval)],
                       new_ds$steps)
```

4. Make a histogram of the total number of steps taken each day and Calculate 
and report the mean and median total number of steps taken per day. Do these 
values differ from the estimates from the first part of the assignment? 
What is the impact of imputing missing data on the estimates of the total daily 
number of steps?

Similarly to the original dataset, in order to create the histogram, the total
daily steps would need to be calculated first. Take note that the option of 
na.rm was not set this time. As we had carried out the substitution in earlier
steps, this setting would be redundant.

Histogram with estimation of any missing data
```{r}
new_dailysteps <- aggregate(steps ~ date, data=new_ds, sum)
hist(new_dailysteps$steps, 
     main="Histogram of the total number of steps taken each day"
     , xlab = "Number of steps daily", ylab = "Number of Days (Frequency)"
     , col="#0000FF88")
```
     
Mean
```{r}
mean(new_dailysteps$steps)
```
Both old dataset where missing data are ignored and new dataset are giving 
the same value of 10766.19 for mean. No differences observed on the mean of 
these two dataset

Median
```{r}
median(new_dailysteps$steps)
```
Median value calculated from old dataset where missing data are ignored is 10765
while new dataset are having median value of 10766.19. Slightly higher value of
median in the new dataset.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” 
and “weekend” indicating whether a given date is a weekday or weekend day.

For this purpose. the library timeDate would be loaded in order to utilised the 
isWeekday() function. 
```{r warning=FALSE}
library(timeDate)
```

The function mutate() from plyr library was used here to create the new calumn.
The new columns was populated using the isWeekday() function which would return
TRUE for weekdays and FALSE for weekend. The columns was then convered to factor
for further use.

```{r}
library(plyr)
new_ds <- mutate(new_ds, WeekDay 
                 = ifelse(
                     isWeekday(as.POSIXct(new_ds$date,origin = "1960-01-01"))
                     ,"weekday" , "weekend"))
new_ds$WeekDay <- as.factor(new_ds$WeekDay)      
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged 
across all weekday days or weekend days (y-axis). The plot should look something
like the following, which was creating using simulated data:

In order to create the graph, an mean of number of steps are first made
on the new datas aggregated by Weekday and interval 
```{r}
weekday_avg <- aggregate(new_ds$steps, 
                         list(interval = new_ds$interval,
                              WeekDay = new_ds$WeekDay), data=new_ds, 
                         FUN=mean)
names(weekday_avg)[3] <- "steps"
```

For the purpose of creating this plot, the lattice library is used and need to 
be loaded before xyplot.
```{r}
library(lattice)
xyplot(weekday_avg$steps~weekday_avg$interval|weekday_avg$WeekDay, 
       scales=list(cex=.8, col="black"),
       xlab="Interval", ylab="Number of Steps", 
       main="Average Number of Steps vs. Interval",
       type = "l",
       layout=(c(1,2)))
```

During weekends, the average number of steps gradually increase from 540 interval
until about 835 interval. The average number of steps continue to fluctuate upto
150 from here onwards.

In contrast, number steps on weekdays gradually increase from 540 interval.
However between 800+, a drastic increase in average steps and it peaked at 835
interval with 230 steps. The average steps decreased from here onwards and from 
900+ onwards - only fluctuating between upto 120.
